Specialized Cost Function for Maximizing Body Area Network Lifetime Using Global Routing Algorithms

### 2011-09-14 docs/thesis/paper.tex
More changes to thesis.tex

\begin{abstract}
$\ll$ at the end $\gg$
\end{abstract}

@@

$\ll$”In this contribution we …. (elaborate on title), describe the algorithm (“in contradistinction to previous routing approached, we define a cost function based on the deviation of the accumulated energy of a single node from the overall average power…”), summary of work (real RSSI data + simulation and hardware implementation using TI 2.4, comparison approach, network lifetime and energy per bit…), summary of results.”$\gg$

@@

\subsection{Link Cost}
The actual link cost is computed by calculating the energy used by the device, \emph{if} that link is selected, and multiplying by a cost factor(Equation~\ref{eqn:link_cost}). The cost factor calculation takes the energy used by the destination device, $E^k_i$, and divides it by the current energy minimum, $E^{min}_{i}$, which is the energy used by the device with the minimum energy used by round $i$. This ratio is then raised to the power $M$, which determines how strong the effect will be $\ll$Need a better way of saying this..$\gg$. If a device's energy is much greater than the current minimum, it will be avoided as a relay.
The $M$ factor in Equation~\ref{eqn:link_cost} can be any number in $0 < M < \infty$. The cost factor is normalized so when $M=0$, it returns to the traditional cost of required power to meet the link.

@@

$\ll$”in what follows we …. (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$

\subsection{Algorithm Implementation}
Detailed pseudo-code for the routing algorithm can be found in Algorithm~\ref{alg:dijkstra} in Appendix~\ref{appendix:algorithms}. Similarly, code for energy used calculation can be found in Algorithm~\ref{alg:energyused}.

\section{Simulation}
$\ll$describe experimental setups and environment (2 setup – walking in room and sitting in from of desk), figure 2 (residential environment for setups)$\gg$

\section{Hardware Implementation}
\subsection{Hardware Platform}
The hardware platform used for this test consists of the Texas Instruments (TI) EZ430-RF2500. This device includes both an MSP430F2274 microcontroller along with a CC2500 2.4GHz transceiver. The CC2500 is configured to run at 250kbps.
A single device, labeled the Access Point (AP), is connected via a USB to Serial link, running at 115200 BAUD, to the host computer. All other devices, labeled End Devices (EDs), are battery powered. In this implementation, the AP acts only as a bridge between the host and the end devices. All routing and power control calculations are done on the host.

@@

Two experimental setups were used. For the first, the subject remained relatively static in a sitting position. The second consisted of a more dynamic environment, where the subject walked around a room as seen in Figure~\ref{fig:floorplan}. The target RSSI ($RSSI^T$) was arbitrarily chosen to be $-60 dBm$.

@@

\subsection{Cycle of Operations}
A normal cycle of operations with $n$ devices works as follows:
\begin{enumerate}
\item AP sends synchronization beacon which includes routing and power control tables.
\item Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power from each.
\item Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices. (See Table~\ref{tab:rssi} in Appendix~\ref{appendix:tables} for an example)
\item The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
\item The host sends both routing and power tables back to the AP so that a new cycle may begin.
\end{enumerate}

@@

$\ll$describe change from simulation (RSSI from power controlled transmission, address benefits$\gg$

@@

@@

$\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$

$\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$

\section{Conclusion}
$\ll$at the end: different from abstract -  The abstract in past tense + concentrate on meaning of results$\gg$

One improvement that can be done deals with the example from Figure~\ref{fig:walk4-c100}. While the failed ED was not part of the network, it was still taken into account when it came to link cost calculation. A better implementation would not include off-line devices in the calculation.

@@

\appendix
\section{Algorithms}\label{appendix:algorithms}
\begin{algorithm}[!ht]
\caption{Dijkstra's Algorithm with new cost function}
\label{alg:dijkstra}
\begin{algorithmic}
\FOR{$i = 1$ to \# of Nodes} 
	\STATE $node_i.visited \leftarrow 0$
	\STATE $node_i.distance \leftarrow \infty$
	\STATE $node_i.previous \leftarrow node_i$
\ENDFOR
\STATE $node_{AP}.distance \leftarrow 0$
\STATE $energy_{min} = min(node.energy)$
\WHILE{unvisited nodes available}
	\STATE $node_{source} \leftarrow $ node with smallest distance
	\FOR{$i = 1$ to \# of Links}
		\IF{$link_i.source = node_{source}$}
			\STATE $node_{dest} \leftarrow link_i.destination$
			\IF{$node_{dest}.visited = 0$}
				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
				\STATE $newdistance \leftarrow node_{source}.distance + cost$
				\IF{$distance < node_{dest}.distance$} 
				\STATE$node_{dest}.distance \leftarrow newdistance$
				\STATE$node_{dest}.previous \leftarrow node_{source}$
				\ENDIF
			\ENDIF
		\ENDIF
	\ENDFOR
	\STATE $node_{source}.visited \leftarrow 1$
\ENDWHILE
\end{algorithmic}
\end{algorithm}

\begin{algorithm}[!ht]
\caption{Energy Used Calculation}
\label{alg:energyused}
\begin{algorithmic}
\FOR{$i = 1$ to \# of Nodes}
	\STATE $node_{ptr} = node_i$
	\IF{$node_{ptr}.previous = node_{ptr}$}
		\STATE No path to $node_i$
	\ELSE
		\WHILE{$node_{ptr}.previous \neq node_{ptr}$}
			\STATE $link_{ptr} \leftarrow $ link between $node_{ptr}$ and $node_{ptr}.previous$
			\STATE $node_{ptr}.energy += link_{ptr}.power$
			\STATE $node_{ptr} \leftarrow = node_{ptr}.previous$
		\ENDWHILE
	\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

\section{Tables}\label{appendix:tables}
\begin{table}[htb]
\label{tab:rssi}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
\hline 
\end{tabular} 
\caption{Sample RSSI Table}
\end{table}

>>>>

\begin{abstract}
TODO
\end{abstract}

@@

In order to provide accurate channel conditions, RSSI measurements of an eight ED network were used as input to our simulations(\ref{sec:simulation}). Real-time hardware implementations were also run to demonstrate the algorithm(\ref{sec:hardware}). The new cost function succeeded in improving network lifetimes when run in a dynamic environment. Static environments did not produce successful results.

@@

\subsection{Link Cost}
The actual link cost is computed by calculating the energy used by the device, \emph{if} that link is selected, and multiplying by a cost factor(Equation~\ref{eqn:link_cost}). The cost factor calculation takes the energy used by the destination device, $E^k_i$, and divides it by the current energy minimum, $E^{min}_{i}$, which is the energy used by the device with the minimum energy used by round $i$. This ratio is then raised to the power $M$, which determines how strong the effect will be. If a device's energy is much greater than the current minimum, it will be avoided as a relay.
The $M$ factor in Equation~\ref{eqn:link_cost} can be any number in $0 < M < \infty$. The cost factor is normalized so when $M=0$, it returns to the traditional cost of required power to meet the link.

@@

\subsection{Algorithm Implementation}
Detailed pseudo-code for the routing algorithm can be found in Algorithm~\ref{alg:dijkstra}. Similarly, code for energy used calculation can be found in Algorithm~\ref{alg:energyused}.

\begin{algorithm}[!ht]
\caption{Dijkstra's Algorithm with new cost function}
\label{alg:dijkstra}
\begin{algorithmic}[1]
\FOR{$i = 1$ to \# of Nodes} 
	\STATE $node_i.visited \leftarrow 0$
	\STATE $node_i.distance \leftarrow \infty$
	\STATE $node_i.previous \leftarrow node_i$
\ENDFOR
\STATE $node_{AP}.distance \leftarrow 0$
\STATE $energy_{min} = min(node.energy)$
\WHILE{unvisited nodes available}
	\STATE $node_{source} \leftarrow $ node with smallest distance
	\FOR{$i = 1$ to \# of Links}
		\IF{$link_i.source = node_{source}$}
			\STATE $node_{dest} \leftarrow link_i.destination$
			\IF{$node_{dest}.visited = 0$}
				\STATE $cost \leftarrow link_i.power * (1 + node_{dest}.energy/energy_{min} )^M/2$
				\STATE $newdistance \leftarrow node_{source}.distance + cost$
				\IF{$distance < node_{dest}.distance$} 
				\STATE$node_{dest}.distance \leftarrow newdistance$
				\STATE$node_{dest}.previous \leftarrow node_{source}$
				\ENDIF
			\ENDIF
		\ENDIF
	\ENDFOR
	\STATE $node_{source}.visited \leftarrow 1$
\ENDWHILE
\end{algorithmic}
\end{algorithm}

\begin{algorithm}[!ht]
\caption{Energy Used Calculation}
\label{alg:energyused}
\begin{algorithmic}[1]
\FOR{$i = 1$ to \# of Nodes}
	\STATE $node_{ptr} = node_i$
	\IF{$node_{ptr}.previous = node_{ptr}$}
		\STATE No path to $node_i$
	\ELSE
		\WHILE{$node_{ptr}.previous \neq node_{ptr}$}
			\STATE $link_{ptr} \leftarrow $ link between $node_{ptr}$ and $node_{ptr}.previous$
			\STATE $node_{ptr}.energy += link_{ptr}.power$
			\STATE $node_{ptr} \leftarrow = node_{ptr}.previous$
		\ENDWHILE
	\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

\section{Simulation}\label{sec:simulation}
In order to test the new cost function in an efficient manner, a simulation was used. Measurements of real channel data were acquired by running the procedure described in~\ref{subsec:cycleofoperations} without the routing or power control enabled. Each ED continuously sampled RSSI data from all other devices and transmitted it to the host.

Two experimental setups were used to capture RSSI data. For the first, the subject remained relatively static in a sitting position. The second consisted of a more dynamic environment, where the subject walked around a room as seen in Figure~\ref{fig:floorplan}.

A set of captured RSSI tables (as seen in Table~\ref{tab:rssi}) were fed into the simulation, which proceeded to run the routing algorithm. The simulation produced energy use data, along with power settings, and routes taken for all devices.

\section{Hardware Implementation}\label{sec:hardware}
In order to verify the new cost function in a real environment, a hardware platform was developed and used. The system function in a similar fashion to the simulation, except that the actual routing and power control data was fed back to the ED's.

\subsection{Hardware Platform}
The hardware platform used for this test consists of the Texas Instruments (TI) EZ430-RF2500. This device includes both an MSP430F2274 microcontroller along with a CC2500 2.4GHz transceiver. The CC2500 is configured to run at 250kbps.
A single device, labeled the Access Point (AP), is connected via a USB to Serial link, running at 115200 BAUD, to the host computer. All other devices, labeled End Devices (EDs), are battery powered. In this implementation, the AP acts only as a bridge between the host and the end devices. All routing and power control calculations are done on the host.

@@

For both simulation and hardware experiments, the target RSSI ($RSSI^T$) was arbitrarily chosen to be $-60 dBm$.

@@

\subsection{Cycle of Operations}\label{subsec:cycleofoperations}
A normal cycle of operations with $n$ devices works as follows:

\begin{algorithmic}[1]
\STATE AP sends synchronization beacon which includes routing and power control tables.
\STATE Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power from each.
\STATE Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices. (See Table~\ref{tab:rssi} for an example)
\STATE The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
\STATE The host sends both routing and power tables back to the AP so that a new cycle may begin.
\end{algorithmic}

\begin{table}[htb]
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline  & AP & ED1 & ED2 & ED3 & ED4 & ED5 & ED6 & ED7 & ED8 \\ 
\hline AP & - & -60 & -60 & -60 & -60.5 & -60.5 & -60.5 & -60 & -79.5 \\ 
\hline ED1 & -66 & - & -62 & -75 & -63.5 & -69 & -75.5 & -73 & -67.5 \\ 
\hline ED2 & -72 & -61.5 & - & -80 & -69 & -70 & -74 & -75 & -66 \\ 
\hline ED3 & -77.5 & -74.5 & -80.5 & - & -67.5 & -70 & -72.5 & -70.5 & -70 \\ 
\hline ED4 & -39 & -63 & -69.5 & -67 & - & -82.5 & -74.5 & -69 & -65.5 \\ 
\hline ED5 & -67.5 & -68 & -71 & -69.5 & -84 & - & -55.5 & -51.5 & -62 \\ 
\hline ED6 & -67 & -75 & -75 & -72.5 & -75 & -55.5 & - & -44.5 & -53 \\ 
\hline ED7 & -46 & -70 & -77.5 & -70.5 & -69 & -50.5 & -45 & - & -51.5 \\ 
\hline ED8 & -78 & -67.5 & -66.5 & -71 & -66 & -62 & -54 & -51.5 & - \\ 
\hline 
\end{tabular} 
\caption{Sample RSSI Table}
\label{tab:rssi}
\end{table}

@@

@@

Due to the similarities between the simulation and hardware implementation (they both use real RSSI data), the results were fairly similar. If fed with the same RSSI data, the algorithm will always produce the same results. The only difference with the hardware implementation is that the RSSI table is reduced due to power control. Since each ED only uses enough transmit power to reach a specific ED, others might not get the message. This was not a problem when the hardware experiment was run.

@@

\section{Conclusion}
After analyzing the results, the new link cost function successfully improves network lifetime. The new cost function outperformed the traditional one in trials dealing with a dynamic environment, where the subject was walking. When the subject was not moving, both link costs performed about the same. Note that, in this case, performance is with respect to network lifetime, and not overall power consumption.
While the new link cost was successful, there is room for improvement. One improvement that can be done deals with the example from Figure~\ref{fig:walk4-c100}. While the failed ED was not part of the network, it was still taken into account when it came to link cost calculation. A better implementation would not include off-line devices in the calculation. A second modification would relate actual battery, or other power source, charge to the algorithm. This would allow for devices to join and leave the network while maintaining proper operation. It would also allow devices with different power sources to be included and used.

@@
